Modelling the innovation and extinction of archaeological ideas

### 2018-03-22 analysis/paper/citation-clusters/How-to-cluster-citations.Rmd
update plots after clustering similar citations

set.seed(1234)
title_word_pairs %>%
  filter(n >= 25) %>%
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(edge_alpha = n, edge_width = n), edge_colour = "cyan4") +
  geom_node_point(size = 5) +
  geom_node_text(aes(label = name), repel = TRUE,
                point.padding = unit(0.2, "lines")) +
  theme_void()
```

@@

title_word_pairs %>%
  filter(n >= 25) %>%
  graph_from_data_frame() %>%
  as_tbl_graph() %>%
  mutate(community = as.factor(group_edge_betweenness())) %>%
  ggraph(layout = 'kk') +
  geom_edge_link(aes(alpha = ..index..), show.legend = FALSE) +
  geom_node_point(aes(colour = community), size = 7) +
    geom_node_text(aes(label = name), repel = TRUE,
                point.padding = unit(0.2, "lines")) +
  theme_graph()

@@

```{r}
citations_community <- 
  title_word_pairs %>% 
 filter(n >= 15) %>%
  graph_from_data_frame() %>%
  as_tbl_graph() %>%
  mutate(community = as.factor(group_edge_betweenness()))  %>% 
  as_tibble %>% 
  arrange(community)

@@

@@

communities_per_year %>% 
  select(time_unit, Species) %>% 
  group_by(time_unit) %>% 
  tally() %>% 
  ggplot(aes(time_unit,
             n)) +
  geom_col() +
  theme_minimal() +
  ggtitle("Number of citation communities per year")

@@

# plot communities over time. 
citation_communities_over_time <- 
communities_per_year %>% 
  filter(Species %in% communities_per_year_wide_range$Species) %>% 
  ggplot(aes(time_unit, 
           perc,
           colour = Species)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_y_log10() +
  theme_minimal() +
  ggtitle("Change over time in citation communities in archaeological literature") +
  xlab("Year") +
  ylab("Percentage of all cited items in that year")

plotly::ggplotly(citation_communities_over_time)
```

@@

hist(xx_with_communities_occ_format_multi_year_hist$number_of_years_cited,
     main = paste("Number of years each article in a citation community was cited for"),
     xlab = "Number of years cited")
```

>>>>

set.seed(1234)

n_cooccurrance <- 10
title_word_pairs_to_plot <- 
title_word_pairs %>%
  mutate(item1 = str_trunc(item1, 30, "right"),
         item2 = str_trunc(item2, 30, "right")) %>% 
  filter(n >= n_cooccurrance) # Erik using 25 from  https://github.com/benmarwick/modelextinctionideas/commit/a530f3e9166c03d89249dfb3621c36e06d2a5233 but a smaller number may be more interesting

title_word_pairs_to_plot %>% 
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(edge_alpha = n, edge_width = n), edge_colour = "cyan4") +
  geom_node_point(size = 5) +
  geom_node_text(aes(label = name), repel = TRUE,
                point.padding = unit(0.2, "lines"),
                   size = 2) +
  theme_void()

ggsave(here::here('analysis/figures/graph_plot1.png'),
       h = 30,
       w = 30)
```

@@

title_word_pairs_to_plot %>% 
  graph_from_data_frame() %>%
  as_tbl_graph() %>%
  mutate(community = as.factor(group_edge_betweenness())) %>%
  ggraph(layout = 'fr') +
  geom_edge_link(aes(alpha = ..index..), 
                 show.legend = FALSE) +
  geom_node_point(aes(colour = community), 
                  size = 2) +
    geom_node_text(aes(label = name), 
                   repel = TRUE,
                   point.padding = unit(0.2, "lines"),
                   size = 0.7) +
  theme_graph()

ggsave(here::here('analysis/figures/graph_plot2.png'),
       h = 30,
       w = 30)

@@

```{r}
n_cooccurrance <- 15 
citations_community <- 
  title_word_pairs %>% 
  filter(n >= n_cooccurrance) %>% 
  graph_from_data_frame() %>%
  as_tbl_graph() %>%
  mutate(community = str_glue('community {as.factor(group_edge_betweenness())}'))  %>% 
  as_tibble %>% 
  arrange(community)

@@

# who are the main authors in each community?
# get first author
main_authors_in_each_community <- 
explore_citation_communities %>% 
  mutate(first_author = str_extract(.out, "(\\w+)")) %>% 
  group_by(Species, first_author) %>% 
  tally() %>% 
  mutate(number = parse_number(Species)) %>% 
  arrange(number, desc(n)) %>% 
  summarise(key_authors = glue::collapse(first_author, sep = ", "),
            number = unique(number)) %>% 
  mutate(AuthorsTitleCase = str_to_title(key_authors)) %>% 
  arrange(number) %>% 
  select(Species, AuthorsTitleCase)

@@

communities_per_year_plot <- 
communities_per_year %>% 
  select(time_unit, Species) %>% 
  group_by(time_unit) %>% 
  tally() %>% 
  ggplot(aes(time_unit,
             n)) +
  geom_col() +
  theme_bw() +
  xlab("") +
  ylab("Number of citation\ncommunities per year")

@@

# plot communities over time. 
library(ggrepel)
communities_per_year_wide_range_subset <- 
communities_per_year %>% 
  filter(Species %in% communities_per_year_wide_range$Species)

# who are in this plot?
main_authors_in_each_community_in_plot <- 
main_authors_in_each_community %>% 
  filter(Species %in% communities_per_year_wide_range_subset$Species)

write_csv(main_authors_in_each_community_in_plot, 
          here::here("analysis/data/derived_data/main_authors_in_each_community_in_plot.csv"))

# now hand edit to assign topics to these communities...
community_2_topics <- readxl::read_excel(here::here('analysis/data/derived_data/main_authors_in_each_community_in_plot.xlsx'))

communities_per_year_wide_range_subset <- 
communities_per_year_wide_range_subset %>% 
  mutate(Species = as.character(Species)) %>% 
  left_join(community_2_topics, by = c('Species' = "Citation community"))

min_y <- min(communities_per_year_wide_range_subset$time_unit)
max_y <- 2027 # give room for line labels

ggplot(communities_per_year_wide_range_subset, 
         aes(time_unit, 
           perc,
           colour = Species)) +
  geom_point(size = 2) +
  stat_smooth(se = FALSE, size = 2) +
  scale_y_log10() +
  theme_minimal(base_size = 15) + 
  geom_text_repel (
    data = subset(communities_per_year_wide_range_subset, 
                  time_unit == max(time_unit)),
    aes(label = Topics),
    size = 6,
    nudge_x = 0,
    hjust = 0,
    segment.color = "grey80",
    direction = "y"
   ) +
   scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017, rep("", length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  theme(legend.position="none",
         panel.grid.minor.x = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  # ggtitle("Change over time in citation communities in the archaeological literature") +
  xlab("Year") +
  ylab("Percentage of all cited items in that year") +
  annotation_custom(grob = ggplotGrob(communities_per_year_plot),
                    xmin = 1975,
                    xmax = 1995,
                    ymin = -1.25,
                    ymax = -0.25)

ggsave(here::here('analysis/figures/citation_communities_over_time.png'),
       h = 10,
       w = 27)

# plotly::ggplotly(citation_communities_over_time)

@@

hist(xx_with_communities_occ_format_multi_year_hist$number_of_years_cited,
     main = paste("Number of years each article in a citation community was cited for"),
     xlab = "Number of years cited")

### 2018-03-22 analysis/paper/paper.Rmd
thinking about how to cluster the citations into taxa
compute on citation clusters to get initial pyrate output
update plots after clustering similar citations

```{r load-data}

@@

# reload --------------------------------------------------------------------
# this is the fast way to resume:
items_df <- readRDS(paste0(here::here(),"/analysis/data/derived_data/saa2018-data-df.rds"))

@@

# get data frame with refs for each article
# each row is a single citation,
# with info of article that has that citation
# the cited article is in the .out column
# xx <-
# items_wth_refs %>%
#   by_row(..f = function(this_row){
#    unlist(str_split(this_row$refs, "\n"))
#   }) %>%
#   unnest(.out) %>%
#   # remove non-refs
#   filter(!grepl("^[A-Z0-9]{2} .*|^ER$", .out)) %>%
#   filter( .out != "") %>%
#   mutate(.out = trimws(toupper(.out))) %>%
#   mutate(.out = str_replace_all(.out, "[[:punct:]]", ""))
# 
# # extract year of cited work
# xx1 <- 
# xx %>% 
#   mutate(cited_year = as.numeric(str_extract(.out, "[0-9]{4}"))) %>% 
#   mutate(cited_year = ifelse(cited_year %in% 1900:2017, 
#                              cited_year,
#                              NA))

# make sure that a cited item has a year that is earlier than the year of the
# item that cited it. Because sometimes there are in press items with the following
# year, we +1 
# xx2 <- 
#   xx1 %>% 
#   filter(cited_year <= year + 1)

# # # # save this as rds to save time
#  saveRDS(xx2,
#            paste0(here::here(), "/analysis/data/derived_data/items_wth_refs.rds"))

# If we are resuming the analysis we can just start from here, and read in the data
# frame of citations.
xx <- readRDS( paste0(here::here(), "/analysis/data/derived_data/items_wth_refs.rds"))
```

@@

We have `r prettyNum(hm_items, big.mark = ",")` articles in our study. The publication dates range from `r time_range[1]` to  `r time_range[2]`. The bibliographic data for these articles was obtained from a search of Clarivate Analytics Web of Science, using the keyword 'archaeolog' and filtering the results to keep only research articles.

@@

@@

Here are the top ten most cited items overall:

```{r}
most_cited <- 
  xx %>% 
  filter(year >= 1975) %>% 
  count(.out, sort = TRUE)  %>% 
  slice(1:10)

@@

pyrate_plot(path_to_python = "C:/python27/python.exe",
            input_file = "pyrate_mcmc_logs/lineage_pyrate_data_0_BD_marginal_rates.log")

>>>>

```{r load-data, eval = FALSE}

@@

# ingest ---------------------------------

# read in the data, TOPIC = "Archaeology", document type = "Article", all years
saa2018_files <- dir("../data/raw_data/wos/",  full.names = TRUE)

text <- map(saa2018_files, ~read_file(.x, locale = locale(encoding = "latin1")))

# split on article delimiter provided by WOS
library(stringr)
items <- unlist(str_split(text, pattern = "\nPT J\n"))
# get rid of the advertising
items <- str_replace_all(items, "FN Clarivate Analytics Web of Science\nVR 1.0", "")
items <- items[items != ""]

length(items) # each item is one article

# function to automate getting the variables out of each item
extractor <- function(i){

  # debug with
  # i <- items[[3]]

  authors =     gsub(".*AU *(.*?) *\nAF .*", "\\1", i)
  authors_n =   str_count(authors, "\n") + 1
  title =       gsub(".*\nTI *(.*?) *\nSO .*", "\\1", i)
  title_n =     str_count(title, " ") - 1
  journal =     gsub(".*\nSO *(.*?) *\nLA .*", "\\1", i)
  abstract =    gsub(".*\nAB *(.*?) *\nC1 .*", "\\1", i)
  refs =        gsub(".*\nCR *(.*?) *\nNR .*", "\\1", i)
  refs_n =      as.numeric(gsub(".*\nNR *(.*?) *\nTC .*", "\\1", i))
  pages_n =     as.numeric(gsub(".*\nPG *(.*?) *\nWC .*", "\\1", i))
  year =        as.numeric(gsub(".*\nPY *(.*?) *\nVL .*", "\\1", i))

  dplyr::data_frame(
    authors =         authors,
    authors_n =       authors_n,
    title =           title ,
    title_n =         title_n,
    journal =         journal,
    abstract  =       abstract,
    refs =            refs    ,
    refs_n =          refs_n ,
    pages_n =         pages_n,
    year =            year
  )
}

# # for debugging, to find the items that break the fn
# for(i in seq_len(length(items))){
#   extractor(items[i])
#   print(i)
# }

# this will take a few mins
items_df <- map_df(items, ~extractor(.x))

saveRDS(items_df, here::here("analysis/data/SAA2018/saa2018-data-df.rds"))

# reload --------------------------------------------------------------------

# this is the fast way to resume:
items_df <- readRDS(here::here("analysis/data/derived_data/saa2018-data-df.rds"))

@@

# get data frame with refs for each article
# each row is a single citation,
# with info of article that has that citation
# the cited article is in the .out column
xx <-
items_wth_refs %>%
  by_row(..f = function(this_row){
   unlist(str_split(this_row$refs, "\n"))
  }) %>%
  unnest(.out) %>%
  # remove non-refs
  filter(!grepl("^[A-Z0-9]{2} .*|^ER$", .out)) %>%
  filter( .out != "") %>%
  mutate(.out = trimws(toupper(.out))) %>%
  mutate(.out = str_replace_all(.out, "[[:punct:]]", ""))
# 
# # extract year of cited work
xx1 <-
xx %>%
  mutate(cited_year = as.numeric(str_extract(.out, "[0-9]{4}"))) %>%
  mutate(cited_year = ifelse(cited_year %in% 1900:2017,
                             cited_year,
                             NA))

# make sure that a cited item has a year that is earlier than the year of the
# item that cited it. Because sometimes there are in press items with the following
# year, we +1 
xx2 <-
  xx1 %>%
  filter(cited_year <= year + 1)

# OpenRefine methods for merging similar citations ----------------------------

# many citations that are actually the same have slight differences in how they appear
# we cna use OpenRefine methods to identify partial string matches

library(refinr)

x <-  xx2$.out

x_refin <- 
x %>% 
  refinr::key_collision_merge() %>% 
  refinr::n_gram_merge()

# Create df for comparing the original values to the edited values.
inspect_results <- 
  data_frame(original_values = x, 
             edited_values = x_refin) %>% 
  mutate(equal = original_values == edited_values)

only_different <-  inspect_results[!inspect_results$equal, c("original_values", "edited_values")]

# update full set of records of citations with the merged ones
xx3 <- 
xx2 %>% 
  left_join(inspect_results,
            by = c(".out" = "original_values")) %>% 
  distinct(authors, 
           authors_n, 
           title, 
           title_n, 
           journal, 
           refs_n, 
           pages_n, 
           abstract, 
           year, 
           cited_year,
           edited_values) %>% 
  mutate(.out = edited_values)  %>% 
  select(-edited_values)

# # # # save this as rds to save time
saveRDS(xx3, here::here("analysis/data/derived_data/items_wth_refs.rds"))
```

```{r}
# If we are resuming the analysis we can just start from here, and read in the data
# frame of citations.
xx <- readRDS( paste0(here::here(), "/analysis/data/derived_data/items_wth_refs.rds"))
```

@@

# how many unique citations?
how_many_unique_citations<- length(unique(xx$.out))
```

We have `r prettyNum(hm_items, big.mark = ",")` articles in our study. The publication dates range from `r time_range[1]` to  `r time_range[2]`. The bibliographic data for these articles was obtained from a search of Clarivate Analytics Web of Science, using the keyword 'archaeolog' and filtering the results to keep only research articles. We have `r how_many_unique_citations` unique citations.

@@

How many citations per article over time?

```{r}
citations_per_year <- 
xx %>% 
  filter(year >= 1975) %>% 
  group_by(abstract, year) %>% 
  tally() %>%
  ungroup() %>% 
  select(year, n)

mean_citations_per_year <- 
xx %>% 
  filter(year >= 1975) %>% 
  group_by(abstract, year) %>% 
   tally() %>%
  group_by(year) %>% 
  summarise(mean_cites = mean(n, na.rm = TRUE)) %>%
  ungroup()

citations_per_year_line <- 
ggplot() + 
  geom_line(data = mean_citations_per_year,
            aes(year, mean_cites)) +
  theme_bw(base_size = 15) +
  ylab("Mean citations/paper")

ggplot(citations_per_year, 
       aes(year, 
             n)) + 
  geom_boxplot(aes(group = year)) +
  theme_minimal(base_size = 15) +
  ylab("Total citations/paper") +
  annotation_custom(grob = ggplotGrob(citations_per_year_line),
                    xmin = 1975,
                    xmax = 1995,
                    ymin = 300,
                    ymax = 600)

ggsave(filename = here::here("analysis/figures/citations_per_year_boxplot_lineplot.png"),
       h = 7,
       w = 10)

@@

Here are the top most cited items overall:

```{r}
most_cited <- 
  xx %>% 
  filter(year >= 1975) %>% 
  count(.out, sort = TRUE)  %>% 
  slice(1:20)

@@

pyrate_plot(path_to_python = "C:/python27/python.exe",
            input_file = "pyrate_mcmc_logs/lineage_pyrate_data_0_BD_marginal_rates.log")
```
